---
title: "INSPECT-SR: Assessing the Trustworthiness of RCTs"
vignette: >
  %\VignetteIndexEntry{inspect_sr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
subtitle: "A structured workflow for detecting problematic trials in systematic reviews"
author: "Your Name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(bayesma)
```

# Introduction

A systematic review is only as trustworthy as the primary studies it includes. Risk of bias tools assess whether a trial was *conducted* properly — they are not designed to detect whether the reported data are *real*.

**INSPECT-SR** (INveStigating ProblEmatic Clinical Trials in Systematic Reviews; Wilkinson et al., 2025) is a structured 21-item tool for assessing the *trustworthiness* of RCTs included in a systematic review. `bayesma` implements the full workflow in four steps:

1. **`inspect_sr()`** — run the assessment (frequentist or Bayesian)
2. **`inspect_sr_table()`** — inspect per-check details
3. **`inspect_plot()` / `inspect_summary_plot()`** — visualise
4. **`filter_trustworthy()`** — filter for meta-analysis

# The four INSPECT-SR domains

| Domain | Items | Focus |
|--------|-------|-------|
| **D1** | 3 | Post-publication notices (retractions, expressions of concern) |
| **D2** | 5 | Conduct, governance, and transparency |
| **D3** | 2 | Text and publication details |
| **D4** | 11 | Results in the study (statistical anomalies) |

Each domain receives **No concerns**, **Some concerns**, or **Serious concerns**. The overall judgement is the most severe domain. `bayesma` automates checks **4.3** (Carlisle), **4.6** (N consistency), **4.8** (GRIM), and **4.9** (p-value verification); the remaining 17 items are read from the input data frame.

# The input data frame

`inspect_sr()` expects a single data frame with **one row per study**:

| Column(s) | Purpose |
|---|---|
| `study` | Study identifier (required, unique) |
| `n_randomised_int`, `n_randomised_ctrl`, `n_randomised_total`, `n_analysed_int`, `n_analysed_ctrl`, `n_lost_int`, `n_lost_ctrl` | CONSORT flow numbers → check 4.6 |
| `d1_1` … `d4_11` | Judgement for each item. Automated items (`d4_3`, `d4_6`, `d4_8`, `d4_9`) are overwritten |
| `d1_1_comment` … `d4_11_comment` | Optional free-text justification |
| `baseline` | **List-column** of Table 1 data frames → GRIM (4.8) and Carlisle (4.3) |
| `statistics` | **List-column** of reported test statistics → p-value check (4.9) |
| `outcome_estimate`, `outcome_ci_lower`, `outcome_ci_upper`, `outcome_se`, `outcome_log_scale` | Optional primary outcome |

# Worked example

The package ships `inspect_sr_example`, an eight-trial simulated review:

```{r}
#| label: load-data

data(inspect_sr_example)

# One row per study — manual items as columns, variable-length data as list-columns
dplyr::glimpse(dplyr::select(inspect_sr_example, study, d1_1, d2_1, d3_1, d4_1))
```

```{r}
#| label: peek-listcols

# Baseline data for the first study
inspect_sr_example$baseline[[1]]

# Test statistics for a study with two tests
inspect_sr_example$statistics[[2]]
```

The eight trials span a range of profiles:

| Study | Expected |
|-------|----------|
| Callahan, Peek, Clayton, Singer | Clean |
| Underwood | Manual D2 "Some concerns" (late registration) |
| Rosa | Carlisle flags too-perfect baseline balance |
| Doe | Retracted; four GRIM failures |
| Kint | CONSORT arithmetic inconsistency |

## Step 1 — `inspect_sr()`

```{r}
#| label: freq-assess

freq_results <- inspect_sr(inspect_sr_example, studyvar = study)
```

The `studyvar` argument uses tidyeval (unquoted), consistent with `bayesma()` and `bayes_forest()`.

## Step 2 — `inspect_sr_table()`

```{r}
#| label: checks-table

# All automated check details
inspect_sr_table(freq_results)
```

```{r}
#| label: checks-failed

# Only failures
inspect_sr_table(freq_results, only_failed = TRUE)
```


```{r}
#| label: checks-per-study
#| eval: false

# Individual Tables
# E.g. Callaghan study
inspect_sr_table(freq_results, study = "Callaghan (1994)")

```

::: {.panel-tabset}

```{r}
#| results: asis

for (s in unique(attr(freq_results, "data")$study)) {
  cat("\n## ", s, "\n\n")
  print(inspect_sr_table(freq_results, study = s))
}
```

:::


The table is grouped by study with thick separators. Failures are highlighted in red; passes in green.

## Step 3 — Visualise

### Traffic-light plot

```{r}
#| label: traffic-light
#| fig-width: 12
#| fig-height: 6

inspect_plot(
  freq_results,
  add_legend = TRUE,
  title    = "INSPECT-SR Trustworthiness Assessment",
  subtitle = "EEG-guided anaesthesia and postoperative delirium"
)
```

Sort by severity:

```{r}
#| label: sorted-plot
#| fig-width: 10
#| fig-height: 5

inspect_plot(freq_results, sort_studies_by = "overall",
             title = "INSPECT-SR Assessment (sorted by concern level)")
```

### Expanded per-item view

Set `incl_checks = TRUE` to see every INSPECT-SR item (1.1 … 4.11) grouped under domain spanner headers, with a per-domain Overall column and a final study-level Overall. The four Domain 4 cells filled in by the automated pipeline (4.3 Carlisle, 4.6 N-consistency, 4.8 GRIM, 4.9 p-value) appear alongside the manual items.

```{r}
#| label: incl-checks-plot
#| #fig-width: 16
#| #fig-height: 5

inspect_plot(freq_results, incl_checks = TRUE,
             title = "INSPECT-SR per-item view")
```

### Summary bar chart

```{r}
#| label: summary-bar
#| fig-width: 8
#| fig-height: 5

inspect_summary_plot(freq_results,
                     title = "Distribution of Trustworthiness Judgements")
```

## Step 4 — Filter for meta-analysis

```{r}
#| label: filter
#| eval: false

# Primary analysis: exclude Serious concerns
data_primary <- filter_trustworthy(
  my_meta_data, freq_results,
  studyvar = study, exclude = "serious"
)

# Sensitivity: also exclude Some concerns
data_strict <- filter_trustworthy(
  my_meta_data, freq_results,
  studyvar = study, exclude = "some"
)

attr(data_primary, "excluded_studies")
```

# The Bayesian extension

Pass `bayes = TRUE` to get Bayes factors instead of pass/fail:

```{r}
#| label: bayes-assess

bayes_results <- inspect_sr(inspect_sr_example, studyvar = study, bayes = TRUE)
```

The per-check BFs are visible in the table:

```{r}
#| label: bayes-table

inspect_sr_table(bayes_results, only_failed = TRUE)
```

## Sensitivity to the prior

```{r}
#| label: prior-sensitivity

r90 <- inspect_sr(inspect_sr_example, studyvar = study, bayes = TRUE,
                  prior_prob_trustworthy = 0.90, verbose = FALSE)
r75 <- inspect_sr(inspect_sr_example, studyvar = study, bayes = TRUE,
                  prior_prob_trustworthy = 0.75, verbose = FALSE)
r50 <- inspect_sr(inspect_sr_example, studyvar = study, bayes = TRUE,
                  prior_prob_trustworthy = 0.50, verbose = FALSE)

data.frame(
  Study   = r90$Study,
  Post_90 = round(r90$Posterior, 3),
  Post_75 = round(r75$Posterior, 3),
  Post_50 = round(r50$Posterior, 3)
)
```

# The automated checks

## GRIM test (check 4.8)

The mean of $n$ integer observations must equal $k/n$ for some integer $k$. A reported mean that fails this is mathematically impossible.

```{r}
#| label: grim-demo

grim_test(3.45, n = 20)   # consistent
grim_test(3.47, n = 20)   # impossible
grim_test(2.45, n = 90)   # impossible (x.x5 at n=90)
```

Bayesian version:

```{r}
#| label: bayes-grim-demo

bayes_grim_test(2.45, n = 90)
```

## P-value verification (check 4.9)

```{r}
#| label: pvalue-demo

verify_pvalue("chi_sq", statistic = 4.12, df = 1, reported_p = 0.04)
verify_pvalue("chi_sq", statistic = 3.84, df = 1, reported_p = 0.004)
```

## Carlisle's test (check 4.3)

```{r}
#| label: carlisle-demo

carlisle_test(c(0.45, 0.12, 0.78, 0.33, 0.91))
carlisle_test(c(0.93, 0.81, 0.95, 0.94, 0.85, 0.95))
```

## Participant number consistency (check 4.6)

```{r}
#| label: n-demo

check_n_consistency(
  n_randomised_int = 150, n_randomised_ctrl = 150,
  n_analysed_int = 142, n_analysed_ctrl = 138,
  n_lost_int = 5, n_lost_ctrl = 8
)
```

# Reporting

> We assessed the trustworthiness of all included RCTs using INSPECT-SR (Wilkinson et al., 2025). Domain 4 checks were automated using the `bayesma` R package (GRIM test, p-value verification, Carlisle's test, participant number consistency). Domains 1–3 were assessed manually by two independent reviewers. Studies with **Serious concerns** were excluded; **Some concerns** studies were retained in a sensitivity analysis.

# References {.unnumbered}

Brown NJL, Heathers JAJ (2017). The GRIM test. *Social Psychological and Personality Science*, 8(4), 363–369.

Carlisle JB (2017). Data fabrication and other reasons for non-random sampling in 5,087 RCTs. *Anaesthesia*, 72(8), 944–952.

Wilkinson J, Heal C, Flemyng E, et al. (2025). INSPECT-SR: a tool for assessing trustworthiness of randomised controlled trials. *BMJ*, 389, e082938.
